Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-30 paper/paper.tex
latest - 30th - d-day

\begin{proof}
Given a Turing machine and an input string, we shall define a function
for   each  non-terminal   introduced  in   the  proof   of
Lemma~\ref{lemma:grammar-undecidable}.
%% Liveness  analysis  of  $f$  will  result in  not  only  contain  a
%% production  $P$ itself,  but could  also contain  extra productions
%% that necessarily contain the \clazy\ symbol.
For instance, corresponding to the production $\nont{i}{\acar} \rightarrow
\acar \nont{j}{\acar}\bcar$,  consider
the program:

@@

If there more than one productions with the same LHS, then the
corresponding function creates joins them with a $\SIF$.   If there
are $n$ such functions,  then \mainpgm\  creates  a $n$-way
branch  and inserts  a  single call  to a  distinct  function in  each
branch. It can  be verified that the grammar for the function shown
above, includes the productions---$\nont{\mathit{all}}{}$
corresponds to the demand $\sigma_{\mathit{all}}$:
\begin{align*}
\nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
\nont{i}{\acar} &\rightarrow \clazy\nont{i}{\acar}\bcar\nont{\mathit{all}}{} \\
\nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
\acdr\nont{\mathit{all}}{}
\end{align*}

Notice that  each production  in Lemma~\ref{lemma:grammar-undecidable}
had a single non-terminal on the RHS. There are two characteristics of
the grammar produced from liveness-analysis  that are are relevant for
replaying the  earlier proof:  (a) If $\mathsf{S}  \rightarrow \beta_1
\mathsf{S'}  \beta_2$  was a  production  in  the earlier  proof,  the
grammar  generated from  liveness  analysis of the constructed program
will  have the  production
$\mathsf{S}      \rightarrow      \beta_1     \mathsf{S'}      \beta_2
\nont{\mathit{\mathit{all}}}{}$,  and (b)  other  productions are  of the  form
$\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
\nont{\mathit{all}}{}$, where $\beta_1'$ and $\beta_2'$ do not contain
any non-terminals.

It is clear that if the TM given as an instance of the halting problem
accepts the input string, then the earlier derivation can be replayed,
every time replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the  sentential   form   by
$\epsilon$. However,  if the TM  does not accept the  input string,
then  any sentential  form derived  from the  start symbol  would have
exactly one non-terminal, and this non-terminal will be different from
$\nont{\mathit{final}}{c}$.  Thus  the  empty   string  would  not  be
derivable from the grammar.
\end{proof}

@@

\section{Experimental Evaluation}
\label{sec:experiments}
  Our benchmark consists of programs taken from nofib~\cite{nofib} and
  other  similar  functional  programming  resources~\cite{PLT-Scheme,
    gc_bench, huffman-sicp}.   All our  benchmarks were executed  on a
  machine having 8  core Intel(R) Core(TM) i7-4770 3.40GHz  CPU , 8192
  KB L2 cache, 16 GB RAM running 64 bit Ubuntu 14.04.

@@

The statistics related to liveness analysis and DFA generation are shown in
Table~\ref{tab:exp-results}(a). We observe that the analysis requires
a reasonable time except for  \verb@treejoin@ and  \verb@sudoku@. 
The bottleneck in our analysis is the NFA to DFA conversion
(worst-case exponential behaviour). To avoid this 
overhead, we use a caching mechanism in our simulator that can
reuse the DFA from earlier analysis if the program has not changed.

Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
statistics  for RGC  and  LGC.  We  report the  number  of garbage collection  events,
average  number of  cells reclaimed  per garbage collection,  average number  of cells
touched  per garbage collection,  and  total time  to
perform  all collections.   It is  no  surprise that  number of  cells
reclaimed per  garbage collection is  higher (and  consequently the  number of  garbage collections is
lower)  for LGC  based collection.   However, the cost  for LGC is that per garbage collection
execution  time is  more  for  LGC, which  results  in higher  overall
execution time  for garbage  collection, even  with reduced  number of
collections.   However, LGC  run time  is still  competitive (slowdown
within \warning{5X} of RGC for worst case), and is better for 3 benchmarks (best
case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
synthetic benchmark  that allocates complete binary  trees of various
sizes.  These binary trees are never used by the program. As a result,
the benchmark highly favours LGC.  We have included this benchmark for
completeness  only  and   the  numbers  for  it   are  not  considered
representative of practical programs.

@@

Memory  usage graphs  for select  benchmarks are  shown
Figure~\ref{fig:memory-usage},  Column   (a).   In some cases, the
number of  garbage collections  tend to be  very large,
hence  we  show only  a  slice of the original graph  which  is
representative  of  the  behavior for  that  particular
benchmark (Column (b)).  In all the programs we can see
that  the   curve  corresponding  to  LGC   (\LGCLine\ line)
regularly dips below the RGC curve (\RGCLine\ line), the dip
is  specially  visible  for the  benchmarks  where  LGC
outperforms RGC substantially.  The graphs also include the
curve    for   reachable    cells   (\ReachLine,    obtained
approximately  by forcing  RGC to  run at  a very  high
frequency)  and  the  live  cells  for a particular run
(\UseLine, obtained by post processing the heap access
at the end of the program).

@@

\section{Related Work} 
\label{sec:relatedwork}
Most previous work on enhancing  garbage collection with liveness were
in   the  context   of   imperative  languages~\cite{Albert,   Hirzel,
  khedker07heap}.    Due  to  the  presence  of  global  variables  and
mutation, these methods are of  limited practical  utility.

There  have been  several efforts  in improving  space utilization  of
functional programs  by compile-time  analysis.  These  are: rewriting
methods                             such                            as
deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
sharing  analysis   based  reallocation~\cite{jones89compile},  region
based   analysis~\cite{tofte98region},   or   through   insertion   of
compile-time                                                nullifying
statements~\cite{inoue88analysis,lee05static,Hamilton}.            All
compile-time marking approaches rely on an efficient and precise alias
analysis and  cannot provide  significant improvement in  its absence.
An interesting  approach suggested by Hofmann~\cite{HofmannJ03}  is to
annotate the heap usage of  first-order programs through linear types.
The annotations are then used to transform the program to serve memory
allocation requests by re-using allocated memory cells.  However, this
requires the user to write programs in a specific way.

@@

%% \begin{figure}[t]
%% \centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
%%  \caption{Memory  usage and garbage collection pattern of  {\tt
%% lambda}.}
%% \label{fig:memory-usage-lambda} \figrule
%% \end{figure}
\vskip -5mm
\section{Conclusions and Future Work}
\label{sec:conclusion}
%% We have  defined a notion of  liveness on structured
%% data; this generalizes classical liveness and strong
%% liveness.    We  started   with   a  general   fully
%% context-sensitive analysis  which we  proved correct
%% with respect  to a minefield semantics  (this models
%% the  effect  of  garbage  collection  between  every
%% evaluation step).

@@

%% One issue we highlighted  was that while fewer nodes
%% were marked (and  more garbage collected), sometimes
%% cons cells  could be visited and  traversed multiple
%% times  with  different  sets of  liveness  paths  to
%% explore.  Further work  includes improvements to the
%% classical copying  collector to  reduce the  cost of
%% this.
We have  extended liveness-based garbage collection  to lazy languages
and shown its benefit for a set of benchmark programs.  This was done
by defining a liveness analysis of heap objects that we proved correct
with  respect   to  a   non-standard  semantics   that  served   as  a
specification of  liveness.  The result  of the  analysis is a  set of
grammars, whose membership question was  proved to be undecidable. The
grammars are  therefore approximated by  DFAs and used by  the garbage
collector to improve collection.

In addition to collecting evaluated  data, our collector also reclaims
closures.   Because  of this  we  had  to  make modifications  to  the
standard run-time representation of a closure to carry liveness of its
free variables and  periodically and update it  during execution time.
As  expected, our  garbage collector  could reclaim  more garbage  and
reduce memory requirements of programs.  An additional benefit was the
reduction in  execution times for  some programs  in spite of  using a
more expensive collection scheme.

>>>>

\begin{proof}
Given a Turing machine and an input string, we shall define a function
for   each  non-terminal   introduced  in   the  proof   of
Lemma~\ref{lemma:grammar-undecidable}.
%% Liveness  analysis  of  $f$  will  result in  not  only  contain  a
%% production  $P$ itself,  but could  also contain  extra productions
%% that necessarily contain the \clazy\ symbol.
For instance, corresponding to the production $\nont{i}{\acar} \rightarrow
\acar \nont{j}{\acar}\bcar$,
the program is:

@@

The body of the function corresponds  to the RHS of the production. If
there are  more than one  productions with the same  LHS non-terminal,
then the function for the non-terminal joins the program fragments for
the  RHS  with a  $\SIF$.   If  there  are  $n$ such  functions,  then
\mainpgm\ creates  a $n$-way  branch and  inserts a  single call  to a
distinct function in each branch. It  can be verified that the grammar
for      the     function      shown      above,     includes      the
productions shown below. $\nont{\mathit{all}}{}$   corresponds   to  the   demand
$\sigma_{\mathit{all}}$:
\begin{align*}
\nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
\nont{i}{\acar} &\rightarrow \clazy\nont{i}{\acar}\bcar\nont{\mathit{all}}{} \\
\nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
\acdr\nont{\mathit{all}}{}
\end{align*}

Notice that  each production  in Lemma~\ref{lemma:grammar-undecidable}
had a single non-terminal on the RHS. There are two characteristics of
the grammar produced from liveness-analysis  that are are relevant for
replaying the  earlier proof:  (a) If $\mathsf{S}  \rightarrow \beta_1
\mathsf{S'}  \beta_2$  was a  production  in  the earlier  proof,  the
grammar  generated from  liveness  analysis of the constructed program
will  have the  production
$\mathsf{S}      \rightarrow      \beta_1     \mathsf{S'}      \beta_2
\nont{\mathit{\mathit{all}}}{}$,  and (b)  other  productions are  of the  form
$\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
\nont{\mathit{all}}{}$.

It is clear that if the TM given as an instance of the halting problem
accepts the input string, then the earlier derivation can be replayed,
every   time   replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the
sentential form by $\epsilon$. However, if  the TM does not accept the
input string, then every sentential form derived from the start symbol
would    have     at    least    one    non-terminal from the grammar
of Lemma~\ref{lemma:grammar-undecidable} that is different
from $\nont{\mathit{final}}{c}$.   Thus  the  empty  string  would  not  be
derivable from the grammar.
\end{proof}

@@

\section{Experimental Evaluation}
\label{sec:experiments}
  Our benchmark consists of programs taken from nofib~\cite{nofib} and
  other repositories for  functional  programs~\cite{PLT-Scheme,
    gc_bench, huffman-sicp}.   All our  benchmarks were executed  on a
  machine having 8  core Intel(R) Core(TM) i7-4770 3.40GHz  CPU , 8192
  KB L2 cache, 16 GB RAM running 64 bit Ubuntu 14.04.

@@

The statistics  related to  liveness analysis  and DFA  generation are
shown in Table~\ref{tab:exp-results}(a). We  observe that the analysis
of  all  programs  except \verb@treejoin@  and  \verb@sudoku@  require
reasonable time.   The bottleneck in  our analysis  is the NFA  to DFA
conversion with  worst-case exponential behaviour. However,  since the
analysis has  to be  done only  once, and its  results can be cached and
re-used, the time spent in analysis may be considered acceptable.

Table~\ref{tab:exp-results}(b) compares  garbage collection statistics
for RGC and  LGC.  We report the number of  garbage collection events,
average  number of  cells  reclaimed per  garbage collection,  average
number of cells  touched per garbage collection and the  total time to
perform all collections.   It is no surprise that the  number of cells
reclaimed per garbage  collection is higher and the  number of garbage
collections  lower  for  LGC.   The  cost of  LGC  is  higher  garbage
collection time, which increases the  overall execution time even with
reduced number of  collections.  However, the execution time  of LGC is
still comparable for most benchmarks (slowdown within \warning{5X} of
RGC in the worst case) and better  for 3 benchmarks (2X speedup in the
best case).  Note that  \verb@gc_bench@~\cite{gc_bench} is a synthetic
benchmark \warning{in the no-fib suite} that allocates complete binary
trees  of various  sizes that  are  never used  by the  program. As  a
result,  the benchmark  highly favours  LGC.  The  benchmark has  been
included  for  completeness,  and  we  do  not  consider  its  numbers
representative of real programs.

@@

Memory  usage graphs  for select  benchmarks are  shown
Figure~\ref{fig:memory-usage},  Column   (a).   In some cases, the
number of  garbage collections  tend to be  very large,
hence  we  show only  a  slice of the original graph  with 
representative    behavior  (Column (b)).  In all the programs we can see
that  the   curve  corresponding  to  LGC   (\LGCLine\ line)
dips   below the RGC curve 
(\RGCLine\ line) during garbage collection.  The graphs also include the
curve    for   reachable    cells   (\ReachLine)  and live cells
(\UseLine). These were  obtained
 by forcing  RGC to  run at   very  high
frequency. The curve for live cells were obtained by recording heap
access times and post processing the data 
at the end of the program.

@@

\section{Related Work} 
\label{sec:relatedwork}

Most previous work on enhancing  garbage collection with liveness were
in   the  context   of   imperative  languages~\cite{Albert,   Hirzel,
  khedker07heap}.    \warning{Due  to  the  presence  of  global  variables  and
mutation, these methods are of  limited practical  utility.}

There  have been  several efforts  in improving  space utilization  of
functional programs  by compile-time  analysis.  These  are: rewriting
methods                             such                            as
deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
sharing  analysis   based  reallocation~\cite{jones89compile},  region
based   analysis~\cite{tofte98region},   or   through   insertion   of
compile-time                                                nullifying
statements~\cite{inoue88analysis,lee05static,Hamilton}.            All
compile-time marking approaches rely on an efficient and precise alias
analysis and  cannot provide  significant improvement in  its absence.
An interesting  approach suggested in~\cite{HofmannJ03}  is to
annotate the heap usage of  first-order programs through linear types.
The annotations are then used to serve memory
requests through re-allocation.  However, this
requires the user to write programs in a specific way.

@@

%% \begin{figure}[t]
%% \centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
%%  \caption{Memory  usage and garbage collection pattern of  {\tt
%% lambda}.}
%% \label{fig:memory-usage-lambda} \figrule
%% \end{figure}
\vskip -5mm
\section{Future Work and Conclusions}
\label{sec:conclusion}
%% We have  defined a notion of  liveness on structured
%% data; this generalizes classical liveness and strong
%% liveness.    We  started   with   a  general   fully
%% context-sensitive analysis  which we  proved correct
%% with respect  to a minefield semantics  (this models
%% the  effect  of  garbage  collection  between  every
%% evaluation step).

@@

%% One issue we highlighted  was that while fewer nodes
%% were marked (and  more garbage collected), sometimes
%% cons cells  could be visited and  traversed multiple
%% times  with  different  sets of  liveness  paths  to
%% explore.  Further work  includes improvements to the
%% classical copying  collector to  reduce the  cost of
%% this.
We have  extended liveness-based  garbage collection  to lazy
  languages  and shown  its benefit  for a  set of  standard benchmark
  programs.  We  defined a liveness analysis  of programs manipulating
  the  heap and  proved  it  correct with  respect  to a  non-standard
  semantics that served as a specification of liveness.  The result of
  the analysis  is a  set of grammars,  whose membership  question was
  shown to be undecidable. The  grammars are therefore approximated by
  DFAs and used by the garbage collector to improve collection.

In addition to  evaluated data, our collector  also reclaims closures.
For this, we  had to modify the standard run-time  representation of a
closures  to carry  liveness  of its free  variables and  periodically
update the liveness during execution  time.  As expected, our garbage
collector  reclaims more  garbage and  reduces memory  requirement of
programs.  An  additional benefit is  that in  spite of using  a more
expensive collector, the execution times remains comparable in
most cases, and even improves for some programs.
